---
title: "Significance & Power"
author: "Filippo Biscarini"
date: "10/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("dplyr")
library("knitr")
library("qvalue")
library("data.table")
```

## Bonferroni correction

Bonferroni correction:

$$
p\_value \leq \frac{\alpha}{m}
$$

$$
p\_value \cdot m \leq \alpha
$$

```{r cars}
results <- fread("../4.gwas/rice_imputed.raw_PH_GWAS.results")

results <- results %>%
  arrange(P)

results %>%
  head() %>%
  kable()
```

```{r}
p_adjusted <- p.adjust(p = results$P, method = "bonferroni")

p_adjusted %>%
  head() %>%
  kable()
```

```{r}
p <- results$P
m <- length(p)
p_adjusted = (p*m)
p_adjusted %>%
  head() %>%
  kable()
```

## q-values

- **q-values**: proxy for FDR based on the distribution of p-values
- $\pi_0$: proportion of true null hypotheses
- **lfdr**: local FDR (empirical Bayesian posterior probability that the null hypothesis is true, conditional on the observed p-value)

```{r pressure}
results <- results %>%
  arrange(P)

qobj <- qvalue(results$P)

qvalues <- qobj$qvalues
pi0 <- qobj$pi0
lfdr <- qobj$lfdr

kable(head(qvalues))
```


```{r}
summary(qobj)
```


```{r}
hist(qobj)
plot(qobj)
```

```{r}
max(qvalues[qobj$pvalues <= 0.01])
```

## Power of GWAS experiments

- **n** = sample size
- **qsq** = fraction of trait variance explained by the SNP ($q^2$; sometimes indicated as $h^2$ to hint at similarity with heritability)
- **beta** = effect size of the SNP on the trait (in SD units: square of beta is used)
- **maf** = minor allele frequency of the SNP ([0,0.5])
- **het** = heterozygous genotype frequency ($2 \cdot \text{maf}*(1-\text{maf})$) under Hardy-Weinberg Equilibrium. (has to be between 0 and 1, usually between 0 and 0.5)
- **pval** = p-value threshold for significance ($\alpha$) 


```{r}
source("power_calc_functions.R")
```

We now calculate the statistical power of GWAS experiments for different sample sizes and proportions of variance explained by the SNPs:

```{r}
pow = power_n_hsq(n = (1:5)*100, qsq = seq(1,25,5)/100, pval=0.001)
kable(pow)
```

```{r, warning=FALSE, message=FALSE}
power_plot(pow, "n", "q-squared")
```
