---
title: "The G Matrix"
author: "Filippo Biscarini"
date: "10/30/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("../software/gMatrix_0.2.tar.gz", type = "source", repos = NULL)
library("knitr")
library("plotly")
library("gMatrix")
library("ggplot2")
library("data.table")
```

## Building the matrix of genomic relationships

Two methods (out of many possible) to calculate genomic relationships between individuals based on genetic variants data:

- Van Raden
- Astle & Balding

Let's have a look at the functions:


```{r vanraden, results='hold'}
gVanRaden.2
```

```{r astle}
gAstleBalding.2
```

### Preparing the data

```{r rice}
data <- fread("rice_imputed.raw")
```

```{r gmatrix}
X <- data[,-c(1,2)] ## remove non-numeric columns
G <- gMatrix::gVanRaden.2(X) ## building the G-matrix
```

```{r}
kable(G[1:10,1:10])
```

### Visualising the G matrix

Heatmap:

```{r}
heatmap(G,col=rev(heat.colors(75)))
```

Multidimensional scaling plot of the distance matrix:

```{r}
D <- (1-G) # first convert relationships to distances
data.mds <- cmdscale(D, k = 3, eig = TRUE)
coords <- as.data.frame(data.mds$points)

p <- ggplot(coords, aes(V1,V2))
p <- p + geom_point()
p
```

Now we add information on the population (metadata):

```{r}
pheno <- fread("../data/rice_phenotypes.txt")
coords$population <- pheno$population[match(data$IID,pheno$id)]

p <- ggplot(coords, aes(V1,V2))
p <- p + geom_point(aes(colour=population))
p
```

### Astle & Balding

```{r, echo=FALSE}
G <- gMatrix::gAstleBalding.2(X)
```

```{r}
kable(G[1:10,1:10])
```

```{r}
heatmap(G,col=rev(heat.colors(75)))
```

```{r}
D <- (1-G) # first convert relationships to distances
data.mds <- cmdscale(D, k = 3, eig = TRUE)
coords <- as.data.frame(data.mds$points)

pheno <- fread("../data/rice_phenotypes.txt")
coords$population <- pheno$population[match(data$IID,pheno$id)]

p <- ggplot(coords, aes(V1,V2))
p <- p + geom_point(aes(colour=population))
p
```

3-D plot:

```{r 3dplot}
p <- plot_ly(data = coords, 
        x = ~V1, y = ~V2, z = ~V3,
        type = "scatter3d",
        color = ~population) %>%
  add_markers() %>%
  layout(scene = list(xaxis = list(title = 'NMDS1'),
                      yaxis = list(title = 'NMDS2'),
                      zaxis = list(title = 'NMDS3')),
         annotations = list(
           x = 0.005,
           y = 0.01,
           text = 'Treatment',
           xref = 'paper',
           yref = 'paper',
           showarrow = FALSE
         ))

p

```

## Exercise

Build and visualize the matrix of genomic relationships using the dog dataset