---
title: "Genotype imputation with KNN: a demonstration"
author: "Filippo Biscarini"
date: "October 25, 2019"
output: html_document
---

# KNN: K-Nearest Neighbors (for classification)

K-nearest neighbors (KNN) is a classification method that attempts to estimate the conditional distribution of the observations (e.g. vector $Y$) given a set of features (matrix $\mathbf{X}$):

$$
Pr(Y=j|X=x_0) = \frac{1}{K} \sum_{i \in \mathcal{N}_0} I(y_i=j)
$$

Let's look at this: 

- $K$ is the chosen **number of neighbors** to consider
- $x_0$ is the **observation** we want to classify (we have features but not know the class it belongs to)
- $\mathcal{N}_0$ is the neighborhood around $x_0$
- $I(y_i=j)$ is the number of neighbors in $\mathcal{N}_0$ belonging to class $j$

The observation $x_0$ is then classified to the class with the largest probability ("Bayes rule").

## Distance function

A crucial aspect in KNN classification is of course how you *define the distance* between observations. A common choice is the Euclidean distance, which basically takes the square root of the sum of **squared differences in feature values** between any two points: 

1. Euclidean distance:

$$
D_E(x_i,x_j) = \sqrt{ \sum_{m=1}^M (x_{i,m}-x_{j,m})^2 }
$$

The Euclidean distance, though, is designed for **continuous features**, and is therefore not very suited for categorical features. Luckily, other distance functions exist for **categorical features**, like the Hamming distance function:

2. Hamming distance (categorical features):

$$
D_H(x_i,x_j) = \frac{1}{M} \sum_{m=1}^M I(x_{i,m} \neq x_{j,m})
$$

In this brief practical/demonstration, we'll go through an illustration of how to apply KNN to the problem of imputing missing SNP genotypes in rice (*Oryza sativa*).

1. Raw data
2. Introduce (artificial) missing genotypes
3. Imputation with R package
4. Step by step KNN imputation: write our own code!

```{r,echo=FALSE,results='hide'}

#Demonstrating the use of KNN to impute missing SNP genotypes
library("DMwR") #for kNN imputation

start.time <- Sys.time()
```

Start time was: `r start.time`

### The dataset

We'll be using SNP genotype data from 12 samples genotyped with the "**GeneChip Rice 44K SNP**".
A summary of the real missing rate in the data is given below:

```{r, echo=FALSE}
## cd ../data
## wget http://www.jackdellequerce.com/data/GenRiz44.txt

rice44 <- read.table("../data/GenRiz44.txt",header=TRUE)
oldcode <- c("AA","BB","--")
neucode <- c(-1,1,NA)

M <- apply(rice44[,5:ncol(rice44)],2,function(x) neucode[match(x,oldcode)])
missingRate <- apply(M,1,function(x) length(x[is.na(x)])/length(x))
summary(missingRate)
M <- M[missingRate==0,]
colnames(M) <- paste("sample",seq(1:ncol(M)),sep="_")
```

All real missing data are edited out of the dataset. This leavs `r nrow(M)` SNP. For the sake of the exercise, initially a much smaller subset with 100 SNPs is extracted.

```{r}
#M <- read.table("rice.dat",header=TRUE)
Mreduced <- t(M[1:100,])
```

Now, we randomly introduce 10\% artificially missing SNP genotypes. In this way, we will be able to measure the accuracy of our imputation method.

```{r}
nSNP <- ncol(Mreduced)
nInd <- nrow(Mreduced)
schwellenWert <- 0.1 # 10% missing data

missing_matrix = matrix(runif(nInd*nSNP), ncol=nSNP) 
missing_matrix[missing_matrix < schwellenWert] = NA
missing_matrix[!is.na(missing_matrix)] = 0
injected_data = Mreduced + missing_matrix

actualMissingRate <- length(injected_data[is.na(injected_data)])/(nInd*nSNP) 

missing_map = missing_matrix
```

The actual missing rate is `r actualMissingRate`.

### Imputation with the R package

We'll now impute missing genotypes using the **DMwR** *R package*. 
In this package there is the *knnImputation()* function, which uses by default the *Euclidean distance*, and allows the user to specify the number of neighbors $K$ and the measure of ``probability'' of belonging to any given class (either **(weighted) average** or **median**).

```{r}
Mimp <- knnImputation(t(injected_data),k=3,meth="median")
Mimp <- t(Mimp)

#Accuracy of imputation
results <- Mimp[is.na(missing_map)]==Mreduced[is.na(missing_map)]
accuracy <- length(results[results==TRUE])/length(results)

```

<span style="background-color:yellow">The accuracy of imputation with the R function *knnImputation()* from the *DMwR* package is:<strong> `r accuracy`</strong></span>

### Step-by-step KNN implementation

Let's have some fun!!

We'll now write our own code. Instead of the Euclidean distance function, we will use the **Hamming distance** function, which is better suited for categorical features. SNP genotypes are categorical features (e.g. **AA/AB/BB**).

First, here's the Hamming distance function:

```{r}
#calculate distance matrix and nearest neighbours
Hamming <- function(ssr_data) {
  
  z <- matrix(0, nrow = nrow(ssr_data), ncol = nrow(ssr_data))
  
  for (k in 1:(nrow(ssr_data) - 1)) {
    for (l in (k + 1):nrow(ssr_data)) {
      z[k, l] <- sum(ssr_data[k, ] != ssr_data[l, ], na.rm=TRUE)
      z[l, k] <- z[k, l]
    }
  }
  
  return(z)
}
```

Again, we choose a small value of $K$ ($K=3$), given the small size of the exercise (and being sure of avoiding the "*curse of dimensionality*" :-)).
The objective is to impute missing SNP genotypes at one SNP locus, using all remaining available information (except the target SNP locus).

```{r,echo=FALSE}
k <- 3

y <- 0
while(length(y[is.na(y)])<2) {
  
  i <- sample(1:ncol(injected_data),1)
  X <- as.matrix(injected_data[,-i]) #global matrix of features (train + test sets)
  y <- injected_data[,i]
}
  
D <- Hamming(X)
#D <- 2-gVanRaden(X)
row.names(D) <- names(y)
colnames(D) <- names(y)
```

Let's have a look at the first elements of the matrix of Hamming distances:

```{r,echo=FALSE}
print(D[1:6,1:6])
```

This is the vector of SNP genotypes to be imputed: `r y`.

Non-missing genotypes at this SNP locus are the "*training*" observations, to be used for the imputation of the missing data points. The subset of distances for the missing observations is extracted and ordered by reverse distance

```{r}
testIDS <- names(y[is.na(y)])
trainIDS <- names(y[!is.na(y)])

NN <- apply(D[as.character(testIDS),as.character(trainIDS)],1,order)
NN <- t(NN)

print(NN)
```

We'll now use KNN to impute the missing genotypes, based on the most frequent genotype in the $K$ neighbors.

```{r}
ergebnisse <- apply(NN[,1:k, drop=F], 1, function(nn) {
    tab <- table(y[trainIDS][nn]); 
    maxClass <- names(which.max(tab))
    prob <- tab[maxClass]/k; 
    pred <- as.integer(maxClass); 
    return(c(pred,prob)) 
})

```

And these are the results for one single SNP locus:

```{r}
ergebnisse <- as.data.frame(t(ergebnisse))
names(ergebnisse) <- c("pred","prob")
ergebnisse[ergebnisse$pred==0,]$prob <- 1-ergebnisse[ergebnisse$pred==0,]$prob
print(ergebnisse)
```

Now we can apply this code to the entire dataset, to impute missing genotypes at all SNP loci.

```{r}
imputedM <- matrix(rep(NA,nInd*nSNP),nrow=nInd)

for(i in 1:ncol(injected_data)) {
  
  X <- as.matrix(injected_data[,-i]) #global matrix of features (train + test sets)
  y <- injected_data[,i]
  k <- 3
  
  if(length(y[is.na(y)])<1) {
    
    imputedM[,i] <- y
  } else {
    
    D <- Hamming(X)
    row.names(D) <- names(y)
    colnames(D) <- names(y)
    
    testIDS <- names(y[is.na(y)])
    trainIDS <- names(y[!is.na(y)])
    
    if(length(testIDS)!=1) {
      
      NN <- apply(D[as.character(testIDS),as.character(trainIDS)],1,order)
      NN <- t(NN)
      ids <- row.names(NN) #for the output file
      
      ergebnisse <- apply(NN[,1:k, drop=F], 1, function(nn) {
        tab <- table(y[trainIDS][nn]); 
        maxClass <- names(which.max(tab))
        pred <- as.integer(maxClass); 
        return(pred)
      })
      y[testIDS] <- ergebnisse[testIDS]
    } else {
      
      n <- order(D[testIDS,trainIDS])[1:k]
      tab <- table(y[trainIDS][n]); 
      maxClass <- names(which.max(tab))
      prob <- tab[maxClass]/k; 
      pred <- as.integer(maxClass); 
      y[testIDS] <- pred
    }

    imputedM[,i] <- y
  }
}

#Accuracy of imputation
results <- imputedM[is.na(missing_map)]==Mreduced[is.na(missing_map)]
accuracy <- length(results[results==TRUE])/length(results)
```

<span style="background-color:yellow">The accuracy of imputation with our R code is:<strong> `r accuracy`</strong></span>

This is expected to be higher than the accuracy obtained with the native *knnImputation()* function, since we used the Hamming distance which is more appropriate for SNP genotype data.

```{r}
end.time <- Sys.time()
time.taken <- end.time - start.time
print("Total time taken below")
print(time.taken)
```

NB: important caveat: the R code shown here is intended for demonstration only, and has not been optimized nor tested for applications.


