---
title: "phenotypes"
author: "Filippo Biscarini"
date: "October 31, 2019"
output: html_document
---

```{r setup, include=FALSE}
library("ggplot2")
library("tidyr")
library("dplyr")
library("knitr")
knitr::opts_chunk$set(echo = TRUE)
```

## Phenotypic data

Exploring phenotypes is equally important as it is exploring genotype data, when approaching a GWAS experiment.
Things to look for are different depending on the type of phenotype e.g. if the trait considered is **continuous** or **binary**.

### Continuous trait

An example is **plant height** (**PH**) from the rice (*Oryza sativa*) dataset.
In a continuos trait, one first looks at summary statistics: average value, standard deviation, minimum and maximum value: 

```{r cars, echo=FALSE}
### Continuous trait
pheno_continuous <- read.table("../data/rice_phenotypes.txt", header = TRUE)
names(pheno_continuous)[3] <- "phenotype"

D <- pheno_continuous %>%
  group_by(population) %>%
  summarise(N=n(), avgPH=mean(phenotype), stdPH=sd(phenotype), minPH=min(phenotype),
            maxPH=max(phenotype))

kable(D)
```

The distribution of a continuos trait is an important aspect: this may give indications as to the best way to model the trait in the GWAS analysis.

We see now a series of density plots: i) overall density plot (all data together)

```{r}
p <- ggplot(pheno_continuous, aes(x=phenotype)) + geom_density()
p <- p + xlab("plant height")
p
```

We add now a vertical line for the average plant height

```{r}
p <- ggplot(pheno_continuous, aes(x=phenotype)) + geom_density()
p <- p + geom_vline(aes(xintercept=mean(phenotype)), color="blue", linetype="dashed", size=1)
p <- p + xlab("plant height")
p
```

Now, we can plot different densities for ech population

```{r}
p <- ggplot(pheno_continuous, aes(x=phenotype)) + geom_density(aes(fill=population), alpha=0.4)
p <- p + scale_fill_manual(values=c("#999999", "#E69F00", "#56B4E9","#FFE4B5"))
p <- p + xlab("plant height")
p
```

And add histograms

```{r}
p <- ggplot(pheno_continuous, aes(x=phenotype, color=population, fill=population))
p <- p + geom_histogram(aes(y=..density..), alpha=0.5, position="identity")
p <- p + geom_density(alpha=.2) + xlab("plant height")
p
```

Boxplots are another effective way to visualize distributions: median, IQR, max, min, outliers

```{r}
p <- ggplot(pheno_continuous, aes(x=population,y=phenotype)) + geom_boxplot(aes(fill=population))
p <- p + xlab("population") + ylab("plant height")
p
```


### Binary trait

Binary (or categorical) traits call for different exploratory statistics.
For instance, the distribution of cases and controls overall and per covariable (when available).
Here we see we have three classes: one is unknown status, which we will remove when filtering phenotypic data before GWAS.

```{r binary, echo=FALSE}
pheno_binary <- read.table("../data/dogs_phenotypes.txt", header = TRUE)

D <- pheno_binary %>%
  group_by(family, phenotype) %>%
  summarise(N=n()) %>%
  spread(key = phenotype, value = N)

kable(D)
```

Barplot of phenotype categories:

```{r pressure, echo=FALSE}
pheno_binary$phenotype <- as.factor(pheno_binary$phenotype)
p <- ggplot(pheno_binary, aes(x=phenotype)) + geom_bar(aes(fill=phenotype))
p <- p + xlab("malformation")
p
```

